COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-25 TACO-paper/coffee.tex
minor fix
Introduction further re-written

%Computational cost is a critical limitation in scientific computing, especially
%for finite element simulations. To provide one particular example we are
%particularly concerned about, it has been well established that mesh resolution
 Consequently, our aggressive optimization of
%local assembly, which may even take up to 60% of the overall FEM's execution
%time, directly impacts the performance of large-scale scientific simulations
%running on supercomputers.

@@

%JUNK: , with synchronization handled at the level of the run-time support

The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of elements, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE. For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts. However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts. In finite elements, numerical kernels are executed during the local assembly phase (or ``assembly'', in the following), which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$. In this paper, we aim at improving the performance of local assembly kernels in a wide variety of notably important PDEs though aggressive code transformation.

@@

Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm. The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates. Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh. As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.

>>>>

%Computational cost is a critical limitation in scientific computing, especially
%for finite element simulations. To provide one particular example we are
%particularly concerned about, it has been well established that mesh resolution
% Consequently, our aggressive optimization of
%local assembly, which may even take up to 60% of the overall FEM's execution
%time, directly impacts the performance of large-scale scientific simulations
%running on supercomputers.

@@

%JUNK: , with synchronization handled at the level of the run-time support
%Computational cost is a critical limitation in finite element simulations. High resolution in time and space is critical to the level of error in the simulations, but computational cost rises superlinearly with resolution. The result of this is that simulations are conducted at the resolution that the user can afford (wall time, allocated supercomputing units or the direct financial cost of cloud time)

The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of cells, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE. For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts. However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts. Hence, the computational cost is a dominant problem in computational science simulations, especially for those based on finite elements, which are the subject of this paper. We address, in particular, the well-known problem of optimizing the local assembly phase of the finite element method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt}, which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$. With respect to these studies, we propose a novel set of composable code transformations targeting, for the first time, instruction-level parallelism - with emphasis on SIMD vectorization - and register locality.

@@

Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm. This kernel can be applied to any element in the mesh: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates. Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh. As already explained, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
